---
title: "Figures 2 and 4 Dose Responses"
output: html_document
---

This code generates the dose response curves in figures 2 and 4 of Stuart et al. 2018.

Load libraries
```{r setup, message = FALSE}
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyr)
library(readr)
```

Define helper functions
```{r helper functions}
plot_facets = function(dose_data, xx, yy, group, label){
ggplot(dose_data, aes_(x = as.name(xx), y = as.name(yy), group = as.name(group)))+
  geom_smooth(method = 'loess', span = 0.5, se=TRUE)+
  scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1))+
  scale_x_log10(labels = fancy_scientific)+
  xlab(label)+
  ylab("Normalized Cell Number")+
  geom_point()+
  theme_bw()+
  facet_wrap(as.name(group))
}
p<-function(f,b) function(a) f(a,b)
find_ic50 = function(y){
  r<-vector("list",2)
  r[[1]] = base::Position(p(`>=`, 0.5), y, right = TRUE)
  r[[2]] = base::Find(p(`>=`,0.5), y, right = TRUE)
  return(r)
}
source("./Helper_Functions/fancy_scientific.R")
```

# Mutants Dose Response to LGX818
```{r LGX818}
lgx_dose = read_csv("../data/lgx_mutant_dose_response.csv")
lgx_dose = gather(lgx_dose, concentration, number, -c(sample, rep_number), convert = TRUE)
factor_levels = as.list(1:7)
names(factor_levels) = unique(lgx_dose$sample)
lgx_dose$sample2 = factor(lgx_dose$sample, levels = unique(lgx_dose$sample))
plot_facets(lgx_dose, "concentration", "number", "sample2", "LGX Concentration (nM)")
ggsave("../results/Dose_Response_Script_Output/lgx_facet.eps", family = "Times")
```

### Figure 2 dose response LGX818
```{r Fig 2 LGX818}
agg_1 = aggregate(lgx_dose$number, by=list(sample2 = lgx_dose$sample2, concentration = lgx_dose$concentration), mean)
agg_2= aggregate(lgx_dose$number, by=list(sample2 = lgx_dose$sample2, concentration = lgx_dose$concentration), sd)

x_axis_labels = c(0.1, 0.3, 1, 3, 10, 30, 100, 300, 1000, "3,000", "10,000")

point_data = data.frame(cbind(agg_1, agg_2$x))
colnames(point_data)[3:4]=c("Mean", "SD")
this_data = dplyr::bind_rows(point_data, lgx_dose)
plot_data = bind_rows(point_data, lgx_dose)
ggplot(plot_data, aes(x =concentration, y = Mean, ymin = (Mean-SD/2), ymax = (Mean+SD/2), group = sample2, color = sample2))+
  geom_point(size = 2)+
  geom_errorbar(width = .2)+
  scale_x_log10(labels = c(0, x_axis_labels), breaks = c(0.01, 0.1, 0.3, 1, 3, 10, 30, 100, 300, 1000, 3000, 10000), minor_breaks = NULL)+
  annotation_logticks(sides = "b", mid = unit(0.1, "cm"))+
  coord_cartesian(xlim = c(0.01, 10000))+
  scale_y_continuous(breaks = c(0, .25, .5, .75, 1))+
  geom_smooth(aes(x = concentration, y = number), method = 'loess', se=FALSE, span = 0.5)+
  labs(color = "sample")+
  xlab("LGX818 Concentration (nM)")+
  ylab("Normalized Cell Number")+
  scale_color_manual(values = c("#000000", "#ff0000", "#3690c0", "#228B22", "#00FFFF", "#FF6600", "#dd3497"))+
  theme_bw()+
  theme(text=element_text(size=14,  family="Times"), axis.text.x = element_text(hjust =1, angle = 45))
ggsave("../results/Dose_Response_Script_Output/lgx_dose_responses.eps", family = "Times")
```

Estimate IC50s.
```{r LGX IC50}
fifty = vector("list", 7)
index = 1
for(ii in unique(lgx_dose$sample)){
  data_subset = filter(lgx_dose, sample == ii)
  fit_line = loess(data_subset$number~log10(data_subset$concentration), span = 0.5)
  start = log10(1)
  end = log10(10000)
  diff = 100
  while(diff >0.0001){
    x = seq(start, end, length.out = 10000)
    y = predict(fit_line, x)
    r = find_ic50(y)
    if(is.na(r[[1]])) break
    start = x[r[[1]]]
    end = x[r[[1]]+1]
    diff = abs(r[[2]]-0.5)
  }
  if(!is.na(r[[1]])) fifty[[index]] = 10^c(x[r[[1]]], x[r[[1]]+1]) %>% mean
  index = index + 1
}
names(fifty) = unique(lgx_dose$sample)
fifty
```

Bootstrap 200 samples to get confidence intervals of IC50
```{r LGX confidence interval}
fifty_ci = vector("list", 7)
confidence_interval = vector("list", 7)
index = 1
for(ii in unique(lgx_dose$sample)){
  val_range = matrix(0, nrow =1, ncol = 200)
  for(jj in 1:200){ii
    data_subset = filter(lgx_dose, sample == ii)
    by_conc = data_subset %>% group_by(concentration)
    boot_data = sample_frac(by_conc, 1, replace = TRUE)
    fit_line = loess(boot_data$number~log10(boot_data$concentration), span = 0.5)
    start = log10(1)
    end = log10(10000)
    diff = 100
    while(diff >0.0001){
      x = seq(start, end, length.out = 10000)
      y = predict(fit_line, x)
      r = find_ic50(y)
      if(is.na(r[[1]])) break
      start = x[r[[1]]]
      end = x[r[[1]]+1]
      diff = abs(r[[2]]-0.5)
    }
    if(!is.na(r[[1]])) val_range[jj] = 10^c(x[r[[1]]], x[r[[1]]+1]) %>% mean
    if(is.na(r[[1]])) break
  }
  fifty_ci[[index]] <- sort(val_range - fifty[[index]])
  confidence_interval[[index]] = fifty[[index]] + fifty_ci[[index]][c(5,195)]
  index = index + 1
}
names(confidence_interval) = unique(lgx_dose$sample)
confidence_interval
```

# Mutants Dose Response to MEK162
```{r MEK162}
mek2_dose = read_csv("../data/mek162_mutant_dose_response.csv")
mek2_dose = gather(mek2_dose, key = concentration, value = number, -c(sample, rep_number), convert = TRUE)
factor_levels = as.list(1:7)
names(factor_levels) = unique(mek2_dose$sample)
mek2_dose$sample2 = factor(mek2_dose$sample, levels = unique(mek2_dose$sample))
plot_facets(mek2_dose, "concentration", "number", "sample2", "MEK162 Concentration (nM)")
ggsave("../results/Dose_Response_Script_Output/mek162_facet.eps", family = "Times")
```

### Generate Figure 3 MEK162 Dose Response
```{r MEK162 Fig 2}
agg_1 = aggregate(mek2_dose$number, by=list(sample2 = mek2_dose$sample2, concentration = mek2_dose$concentration), mean)
agg_2= aggregate(mek2_dose$number, by=list(sample2 = mek2_dose$sample2, concentration = mek2_dose$concentration), sd)

x_axis_labels = c(0.1, 0.3, 1, 3, 10, 30, 100, 300, 1000, "3,000", "10,000")

point_data = data.frame(cbind(agg_1, agg_2$x))
colnames(point_data)[3:4]=c("Mean", "SD")
this_data = dplyr::bind_rows(point_data, mek2_dose)
plot_data = bind_rows(point_data, mek2_dose)
ggplot(plot_data, aes(x =concentration, y = Mean, ymin = (Mean-SD/2), ymax = (Mean+SD/2), group = sample2, color = sample2))+
  geom_point(size = 2)+
  geom_errorbar(width = .2)+
  scale_x_log10(labels = c(0, x_axis_labels), breaks = c(0.01, 0.1, 0.3, 1, 3, 10, 30, 100, 300, 1000, 3000, 10000), minor_breaks = NULL)+
  annotation_logticks(sides = "b", mid = unit(0.1, "cm"))+
  scale_y_continuous( breaks = c(0, .25, .5, .75, 1))+
  geom_smooth(aes(x = concentration, y = number), method = 'loess', se=FALSE, span = 0.75)+
  labs(color = "sample")+
  coord_cartesian(xlim = c(0.01, 10000), ylim = c(0, 1.15))+
  xlab("MEK162 Concentration (nM)")+
  ylab("Normalized Cell Number")+
  scale_color_manual(values = c("#000000", "#ff0000", "#3690c0", "#228B22", "#00FFFF", "#FF6600", "#dd3497"))+
  theme_bw()+
  theme(text=element_text(size=14,  family="Times"), axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("../results/Dose_Response_Script_Output/mek162_dose_responses.eps", family = "Times")
```

Estimate IC50
```{r MEK162 IC50}
fifty = vector("list", 6)
index = 1
for(ii in unique(mek2_dose$sample)){
  data_subset = filter(mek2_dose, sample == ii)
  fit_line = loess(data_subset$number~log10(data_subset$concentration), span = 0.5)
  start = log10(1)
  end = log10(10000)
  diff = 100
  while(diff >0.0001){
    x = seq(start, end, length.out = 10000)
    y = predict(fit_line, x)
    r = find_ic50(y)
    if(is.na(x[r[[1]]+1])) break
    start = x[r[[1]]]
    end = x[r[[1]]+1]
    diff = abs(r[[2]]-0.5)
  }
  if(!is.na(r[[1]])) fifty[[index]] = 10^c(x[r[[1]]], x[r[[1]]+1]) %>% mean
  index = index + 1
}
names(fifty) = unique(mek2_dose$sample)
fifty
```

Bootstrap 200 samples to get confidence intervals of IC50
```{r MEK162 confidence interval}
fifty_ci = vector("list", 6)
confidence_interval = vector("list", 6)
index = 1
for(ii in unique(mek2_dose$sample)){
  val_range = matrix(0, nrow =1, ncol = 200)
  for(jj in 1:200){
    data_subset = filter(mek2_dose, sample == ii)
    by_conc = data_subset %>% group_by(concentration)
    boot_data = sample_frac(by_conc, 1, replace = TRUE)
    fit_line = loess(boot_data$number~log10(boot_data$concentration), span = 0.5)
    start = log10(1)
    end = log10(10000)
    diff = 100
    while(diff >0.0001){
      x = seq(start, end, length.out = 10000)
      y = predict(fit_line, x)
      r = find_ic50(y)
      if(is.na(x[r[[1]]+1])) break
      start = x[r[[1]]]
      end = x[r[[1]]+1]
      diff = abs(r[[2]]-0.5)
    }
    if(!is.na(r[[1]])) val_range[jj] = 10^c(x[r[[1]]], x[r[[1]]+1]) %>% mean
    if(is.na(r[[1]])) break
  }
  fifty_ci[[index]] <- sort(val_range - fifty[[index]])
  confidence_interval[[index]] = fifty[[index]] + fifty_ci[[index]][c(5,195)]
  index = index + 1
}
names(confidence_interval) = unique(mek2_dose$sample)
confidence_interval
```

# Time Course Dose Responses
```{r, message = FALSE}
dose_data = read_csv("../data/intermittent_dose_response.csv")
dose_data2 = gather(dose_data, concentration, number, -c(sample, rep_number), convert = TRUE)
plot_facets(dose_data2, "concentration", "number", "sample", "LGX Concentration (nM)")
ggsave("../results/Dose_Response_Script_Output/timecourse_facet_dose_responses.eps")
```

### Create figure 4 dose responses
#### Continuous Dose Response
```{r continuous, warning = FALSE, message = FALSE}
dose_data_continuous = filter(dose_data2, sample =="Vector Control" | sample =="V600E" | sample =="Week 1" |sample =="Week 2 C" | sample =="Week 3 C" | sample =="Week 4 C")

agg_1 = aggregate(dose_data_continuous$number, by=list(sample = dose_data_continuous$sample, concentration = dose_data_continuous$concentration), mean)
agg_2= aggregate(dose_data_continuous$number, by=list(sample = dose_data_continuous$sample, concentration = dose_data_continuous$concentration), sd)

point_data = data.frame(cbind(agg_1, agg_2$x))
this_data = dplyr::bind_rows(point_data, dose_data_continuous)
colnames(point_data)[3:4]=c("Mean", "SD")

x_axis_labels = c(0.1, 0.3, 1, 3, 10, 30, 100, 300, 1000, "3,000", "10,000")

plot_data = bind_rows(point_data, dose_data_continuous)
ggplot(plot_data, aes(x =concentration, y = Mean, ymin = (Mean-SD/2), ymax = (Mean+SD/2), group = sample, color = sample))+
  geom_point(size = 2)+
  geom_errorbar(width = .2)+
    scale_x_log10(labels = c(0, x_axis_labels), breaks = c(0.01, 0.1, 0.3, 1, 3, 10, 30, 100, 300, 1000, 3000, 10000), minor_breaks = NULL)+
  annotation_logticks(sides = "b", mid = unit(0.1, "cm"))+
  scale_y_continuous(limits = c(0, 1.09), breaks = c(0, .25, .5, .75, 1))+
  geom_smooth(aes(x = concentration, y = number), method = 'loess', se=FALSE, span = 0.5)+
  xlab("LGX818 Concentration (nM)")+
  ylab("Normalized Cell Number")+
  scale_color_manual(values = c("#4c4c4c", "#a9a9a9", "#bae4bc", "#7bccc4", "#43a2ca", "#0868ac"))+
  coord_cartesian(xlim = c(0.01, 10000), ylim = c(0, 1.05))+
  theme(text=element_text(size=14,  family="Times"), axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("../results/Dose_Response_Script_Output/continuous_dose_response.eps", family = "Times")
```

#### Intermittent Dose Response
```{r Intermittent, warning = FALSE, message=FALSE}
dose_data_int = filter(dose_data2, sample =="Vector Control" | sample =="V600E" | sample =="Week 1" |sample =="Week 2 I" | sample =="Week 3 I" | sample =="Week 4 I")

agg_1 = aggregate(dose_data_int$number, by=list(sample = dose_data_int$sample, concentration = dose_data_int$concentration), mean)
agg_2= aggregate(dose_data_int$number, by=list(sample = dose_data_int$sample, concentration = dose_data_int$concentration), sd)

point_data = data.frame(cbind(agg_1, agg_2$x))
this_data = dplyr::bind_rows(point_data, dose_data_int)
colnames(point_data)[3:4]=c("Mean", "SD")

plot_data = bind_rows(point_data, dose_data_int)
INT = data.frame(INT = (substr(plot_data$sample, 8,8)=="I"))
plot_data = bind_cols(plot_data, INT)
ggplot(plot_data, aes(x =concentration, y = Mean, ymin = (Mean-SD/2), ymax = (Mean+SD/2), group = sample, color = sample, linetype = INT))+
  geom_point(size = 2)+
  geom_errorbar(width = .2, linetype = "solid")+
  scale_x_log10(labels = c(0, x_axis_labels), breaks = c(0.01, 0.1, 0.3, 1, 3, 10, 30, 100, 300, 1000, 3000, 10000), minor_breaks = NULL)+
  annotation_logticks(sides = "b", mid = unit(0.1, "cm"))+
  scale_y_continuous(limits = c(0, 1.09), breaks = c(0, .25, .5, .75, 1))+
  geom_smooth(aes(x = concentration, y = number, linetype = INT), method = 'loess', se=FALSE, span = 0.5)+
  xlab("LGX818 Concentration (nM)")+
  ylab("Normalized Cell Number")+
  coord_cartesian(xlim = c(0.01, 10000), ylim = c(0, 1.05))+
  scale_linetype_discrete(guide = FALSE)+
  scale_color_manual(values = c("#4c4c4c", "#a9a9a9", "#bae4bc", "#7bccc4", "#43a2ca", "#0868ac"))+
  theme(text=element_text(size=14,  family="Times"), axis.text.x = element_text(angle = 45, hjust = 1, size = 12))
ggsave("../results/Dose_Response_Script_Output/intermittent_dose_response.eps", family = "Times")
```

Estimate IC50s
```{r IC50 estimate}
fifty = vector("list", 9)
index = 1
for(ii in unique(dose_data2$sample)){
  data_subset = filter(dose_data2, sample == ii)
  fit_line = loess(data_subset$number~log10(data_subset$concentration), span = 0.5)
  start = log10(1)
  end = log10(10000)
  diff = 100
  while(diff >0.0001){

    x = seq(start, end, length.out = 10000)
    y = predict(fit_line, x)
    r = find_ic50(y)
    start = x[r[[1]]]
    end = x[r[[1]]+1]
    diff = abs(r[[2]]-0.5)
  }
  fifty[[index]] = 10^c(x[r[[1]]], x[r[[1]]+1]) %>% mean
  index = index + 1
}
names(fifty) = unique(dose_data$sample)
fifty
```

Bootstrap samples to get confidence intervals for IC50s
```{r Confidence Interval}
fifty_ci = vector("list", 9)
confidence_interval = vector("list", 9)
index = 1
for(ii in unique(dose_data$sample)){
  val_range = matrix(0, nrow =1, ncol = 200)
  for(jj in 1:200){
    data_subset = filter(dose_data2, sample == ii)
    by_conc = data_subset %>% group_by(concentration)
    boot_data = sample_frac(by_conc, 1, replace = TRUE)
    fit_line = loess(boot_data$number~log10(boot_data$concentration), span = 0.5)
    start = log10(1)
    end = log10(10000)
    diff = 100
    while(diff >0.0001){
      x = seq(start, end, length.out = 10000)
      y = predict(fit_line, x)
      r = find_ic50(y)
      start = x[r[[1]]]
      end = x[r[[1]]+1]
      diff = abs(r[[2]]-0.5)
    }
    val_range[jj] = 10^c(x[r[[1]]], x[r[[1]]+1]) %>% mean
  }
  fifty_ci[[index]] <- sort(val_range - fifty[[index]])
  confidence_interval[[index]] = fifty[[index]] + fifty_ci[[index]][c(5,195)]
  index = index + 1
}
names(confidence_interval) = unique(dose_data2$sample)
confidence_interval
```

```{r sessionInfo}
sessionInfo()
```

