---
title: "Resistance and Sensitivity Genes"
output: html_document
---
This code generates the heatmaps of resistance and sensitivity genes in figure 7 and supplemental figure S3 in Stuart et al. 2018.

First load in libraries
```{r libraries, message = FALSE, warning=FALSE}
library(DESeq2)
library(magrittr)
library(dplyr)
library(BiocParallel)
library(readr)
```

Load in preprocessed rna seq data image created by the RNA_SEQ_Script.Rmd file.
```{r helper functions, message = FALSE}
load("../data/processed_rna_seq_data.rda")
source("./Helper_Functions/dend_sort_exp.R")
source("./Helper_Functions/plot_heatmap.R")
source("Helper_Functions/sort_exp.R")
```

Run hypothesis tests for the reversible and irreversible criteria.
```{r hypo tests}
lfc_thresh = 0.97
a_thresh = 0.05

# do these significantly change by an abs(lfc) greater than 0.97?
r24 = results(dds, contrast = c('condition', '2', '4'), alpha = a_thresh, lfcThreshold = lfc_thresh, altHypothesis = "greaterAbs", parallel = TRUE, BPPARAM = MulticoreParam(4))
r26 = results(dds, contrast = c('condition', '2', '6'), alpha = a_thresh, lfcThreshold = lfc_thresh, altHypothesis = "greaterAbs", parallel = TRUE, BPPARAM = MulticoreParam(4))
r25 = results(dds, contrast = c('condition', '2', '5'), alpha = a_thresh, lfcThreshold = lfc_thresh, altHypothesis = "greaterAbs", parallel = TRUE, BPPARAM = MulticoreParam(4))
r27 = results(dds, contrast = c('condition', '2', '7'), alpha = a_thresh, lfcThreshold = lfc_thresh, altHypothesis = "greaterAbs", parallel = TRUE, BPPARAM = MulticoreParam(4))
r29 = results(dds, contrast = c('condition', '2', '9'), alpha = a_thresh, lfcThreshold = lfc_thresh, altHypothesis = "greaterAbs", parallel = TRUE, BPPARAM = MulticoreParam(4))

r28 =  results(dds,contrast = c('condition', '2', '8'),  alpha=a_thresh, lfcThreshold = lfc_thresh, altHypothesis = "greaterAbs", parallel = TRUE, BPPARAM = MulticoreParam(4))
r210 =  results(dds,contrast = c('condition', '2', '10'),  alpha=a_thresh, lfcThreshold = lfc_thresh, altHypothesis = "greaterAbs", parallel = TRUE, BPPARAM = MulticoreParam(4))
r56 = results(dds, contrast = c('condition', '5', '6'), alpha = a_thresh, lfcThreshold = lfc_thresh, altHypothesis = "greaterAbs", parallel = TRUE, BPPARAM = MulticoreParam(4))
r68 =  results(dds,contrast = c('condition', '6', '8'),  alpha=a_thresh, lfcThreshold = lfc_thresh, altHypothesis = "greaterAbs", parallel = TRUE, BPPARAM = MulticoreParam(4))
r108 =  results(dds,contrast = c('condition', '10', '8'),  alpha=a_thresh, lfcThreshold = lfc_thresh, altHypothesis = "greaterAbs", parallel = TRUE, BPPARAM = MulticoreParam(4))

unadjusted_pvals = c(r24$pvalue, r26$pvalue, r56$pvalue, r68$pvalue, r108$pvalue, r25$pvalue, r27$pvalue, r29$pvalue, r28$pvalue, r210$pvalue)
readjust_pvals = unadjusted_pvals %>% p.adjust(method = "fdr")

tests = list(r24 = r24, r26 = r26, r56 = r56, r68 = r68, r108 = r108, r25 = r25, r27 = r27, r29 = r29, r28 = r28, r210 =r210)
len_save = 1
for(ii in names(tests)){
  len = dim(tests[[ii]])[1]+len_save-1
  tests[[ii]]$p_readjust = readjust_pvals[len_save:len]
  len_save = len+1
}
```

Identify genes that meet reversibile and irreversible criteria.
```{r criteria}
meet_reversible_criteria = which(tests[["r56"]]$p_readjust<a_thresh & tests[["r68"]]$p_readjust<a_thresh & tests[["r108"]]$p_readjust<a_thresh
                                 & (sign(tests[["r68"]]$log2FoldChange) ==sign(tests[["r108"]]$log2FoldChange)))
reversible_genes = gene_names[meet_reversible_criteria, ]

meet_irreversible_criteria = which(tests[["r25"]]$p_readjust<a_thresh &
    tests[["r27"]]$p_readjust<a_thresh & tests[["r29"]]$p_readjust<a_thresh & 
    tests[["r26"]]$p_readjust<a_thresh & tests[["r28"]]$p_readjust<a_thresh &
    tests[["r210"]]$p_readjust<a_thresh &
((sign(tests[["r25"]]$log2FoldChange) == sign(tests[["r27"]]$log2FoldChange)) &
(sign(tests[["r25"]]$log2FoldChange) == sign(tests[["r29"]]$log2FoldChange)) &
(sign(tests[["r25"]]$log2FoldChange) == sign(tests[["r26"]]$log2FoldChange)) &
(sign(tests[["r25"]]$log2FoldChange) == sign(tests[["r28"]]$log2FoldChange)) &
(sign(tests[["r25"]]$log2FoldChange) == sign(tests[["r210"]]$log2FoldChange)))) %>% setdiff(meet_reversible_criteria)

irreversible_genes = gene_names[meet_irreversible_criteria, ]

reversible_genes %>% write_csv("../results/Resistance_and_Sensitivity_Genes_Script_Output/reversible_changes_genes.csv")
irreversible_genes %>% write_csv("../results/Resistance_and_Sensitivity_Genes_Script_Output/irreversible_changes_genes.csv")
```

Load in resistance and sensitivity gene names and categories.
```{r load data}
rs_genes = read.csv("../data/resistance_sensitivity_genes.csv", stringsAsFactors = FALSE)
map = match(rs_genes$symbol, gene_names$symbol)
lfc = as.data.frame(rlog_data[map,] - rowMeans(rlog_data[map,c(1:3)]))
colnames(lfc) = c(rep("Vector", 3), rep("V600E", 3), rep("20hr Drug", 3), rep("Week 1 CONT", 3), rep("Week 2 CONT", 3), rep("Week 2 INT (OFF)", 3), rep("Week 3 CONT", 3), rep("Week 3 INT (ON)", 3), rep("Week 4 CONT", 2), rep("Week 4 INT (OFF)", 3))

```

Confirm that all the reversible/irreversible resistance/sensitivity genes meet the above criteria.
```{r confirm conditions}
reversible_rs = !(match(rs_genes$symbol, reversible_genes$symbol) %>% is.na)
rs_genes$symbol[!(reversible_rs ==(rs_genes$category == "RR" |rs_genes$category == "RS"))]
rs_genes$category[!(reversible_rs ==(rs_genes$category == "RR" |rs_genes$category == "RS"))] = "O"
irreversible_rs = !(match(rs_genes$symbol, irreversible_genes$symbol) %>% is.na)
rs_genes$symbol[!(irreversible_rs ==(rs_genes$category == "IC"))]
rs_genes$category[!(irreversible_rs ==(rs_genes$category == "IC"))] = "O"

lfc_data_frame = data.frame(gene_name = gene_names[map,], category = rs_genes$category, lfc)
```

## Figure 7 Heatmap
```{r heatmap 1, message = FALSE}
index = 0
lfc_data_frame2 = data.frame()
categories = c("IC", "RS", "RR")
horz_breaks = array(NA, dim = c(0,0))
for (ii in categories) {
  sub_data = lfc_data_frame %>% filter(category == ii)
  index = index + dim(sub_data)[1]
  horz_breaks = c(horz_breaks, index)
  rownames(sub_data) = sub_data$gene_name.symbol
  dist.mat = dist(sub_data[,8:33])
  hc = hclust(dist.mat, method = "ward.D2")
  mean_exp = round(rowMeans(sub_data[hc$order,c(20:22, 26:28, 31:33)]),4)
  hc2 = dend_sort_exp(hc, mean_exp) #sort the dendrogram
  sub_data = sub_data[hc2$order, ]
  lfc_data_frame2 = bind_rows(lfc_data_frame2, sub_data) #replace in lfc
}
lfc_only = lfc_data_frame2[-c(1:4)]
lfc_only = lfc_only[,c(1:15, 19:21,25,26,16:18,22:24,27:29)]
horz_breaks = horz_breaks[-3]
plot_heatmap(lfc_only, thresh = 0.97, cluster_rows = FALSE, cluster_cols = FALSE, gaps_col = c(9,20), gaps_row = horz_breaks, legend_breaks = c(-6, -4, -2, -0.97, 0, 0.97, 2, 4, 6), labels_row = lfc_data_frame2$gene_name.symbol, show_colnames=FALSE, fontsize = 7)
dev.copy2eps(file="../results/Resistance_and_Sensitivity_Genes_Script_Output/Resistance_and_Sensitivity_Genes_Script_Output.eps")
```

## Supplmental Figure S3
```{r heatmap 2, message = FALSE}
sub_data = lfc_data_frame %>% filter(category == "O")
rownames(sub_data) = sub_data$gene_name.symbol
dist.mat = dist(sub_data[,8:33])
hc = hclust(dist.mat, method = "ward.D2")
mean_exp = round(rowMeans(sub_data[hc$order,8:33]),4)
hc2 = dend_sort_exp(hc, mean_exp) #sort the dendrogram

lfc_other_only = sub_data[-c(1:4)]
lfc_other_only = lfc_other_only[,c(1:15, 19:21,25,26,16:18,22:24,27:29)]

plot_heatmap(lfc_other_only, thresh = 0.97, cluster_rows = hc2, cluster_cols = FALSE, gaps_col = c(9,20), legend_breaks = c(-6, -4, -2, -0.97, 0, 0.97, 2, 4, 6), labels_row = lfc_other_only$gene_name.symbol, show_colnames=FALSE, fontsize = 7)
dev.copy2eps(file="../results/Resistance_and_Sensitivity_Genes_Script_Output/Supplemental_Resistance_and_Sensitivity_Genes_Script_Output.eps")
```


```{r}
sessionInfo()
```

